---
title: "Using UFO Vectors in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using UFO Vectors in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

~~~ 
Note:

This is an R markdown document. It is like a markdown document, but has
executable snippets of code in in R and bash. It can be viewed using vscode or
RStudio, e.g.:

    rstudio ufo-r-vectors.Rmd

There, each chunk can be executed via a button over the code. All paths in this
file are relative to the location of the file, so set the session's working
directory to the source file in rstudio before running the snippets:

    Session -> Set working directory -> To source file location

The file can also be executed (in whole) with R using the following bash
snippet:*
 
    Rscript -e "rmarkdown::render('ufo-r-vectors.Rmd')"

This will compile the document into HTML while executing all the code snippets
in the document. 
~~~

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "# "
)
```

# Overview

Package *ufovectors* is an example implementation of R vectors using Userfault
Objects (UFOs) from the *ufos* package. UFOs allow our R vectors to be lazily
loaded into memory on demand from an arbitrary *source*. UFO vectors can be
larger than memory, with data being transparentl loaded from source on demand
and unloaded as needed.

Package *ufovectors* provides a sample implementations of multiple UFO backends:

```r
# Large from-to-by sequences
ufo_integer_seq   <- function(from, to, by, ...) #
ufo_numeric_seq   <- function(from, to, by, ...) #

# File-backed vectors (a la mmap)
ufo_integer_bin   <- function(path, ...) #
ufo_numeric_bin   <- function(path, ...) #
ufo_complex_bin   <- function(path, ...) #
ufo_logical_bin   <- function(path, ...) #
ufo_raw_bin       <- function(path, ...) #     

# File-backed vectors, but the file is compressed
ufo_integer_bz2   <- function(path, ...) #
ufo_numeric_bz2   <- function(path, ...) #
ufo_complex_bz2   <- function(path, ...) #
ufo_logical_bz2   <- function(path, ...) #
ufo_raw_bz2       <- function(path, ...) #
ufo_character_bz2 <- function(path, ...) #

# Composition of multiple vectors
ufo_bind          <- function(...)                            #

# Write protection wrapper
ufo_write_protect <- function(vector, ...)                    #

# Vector backed by a database column
ufo_psql          <- function(db, table, column, ...)         #

# A data.frame of vectors backed by a CSV file
ufo_csv           <- function(path, check_names, header, ...) #

# Empty larger-than-memory vectors
ufo_integer       <- function(size, populate_with_NAs, ...) #
ufo_numeric       <- function(size, populate_with_NAs, ...) #
ufo_complex       <- function(size, populate_with_NAs, ...) #
ufo_logical       <- function(size, populate_with_NAs, ...) #
ufo_raw           <- function(size,                    ...) #
ufo_character     <- function(size, populate_with_NAs, ...) #
```

All constructors can be configured with a `min_load_count` flag specifying the
chunk size of a UFO in terms of elements, and a `read_only` flag, which turns on
write-protection. The size of the chunk depends must be at least a page of
memory. Example:

```r
ufo_integer_seq(1, 100000, 1, read_only=TRUE, min_load_count=4096)
```

# Installation

The UFO framework comes as two packages: `ufos` and `ufovectors`. The former
contains a general framework for implementing programmable memory objects,
whereas the latter is a library of example implementationsof larger-than-memory
vectors. Install both form source via:

```bash
R CMD INSTALL --preclean ../projects/ufo-r 
R CMD INSTALL --preclean ../projects/ufo-r-vectors
```

# Basic usage

We show basic usage of the UFO framework.

## Loading the library

Once in R, load the UFO vector library into your program:

```{r}
options(ufos.high_water_mark_mb=2, ufos.low_water_mark_mb=1)
library(ufovectors)
```

This also loads the `ufos` and package as a dependency, as well as a `viewports`
package used to do custom subsetting on UFOs.

It is possible to optionally specify a high and low watermark for the framework.
The high water mark is the maximum memory that can be devoted to holding
materialized chunks in memory. The low water mark is the amount of memory the
chunk GC will retrain when it unloads chunks. The values are given in MB and are
set here to 2MB and 1MB, respectively. This is purposefully set low for
demonstration purposes. They are set to 2048MB and 1024MB by default.

When UFOs are loaded, the UFO framework thread is automatically started. When
the library is unloaded or the runtime is closed, the UFO runtime cleans up
after itself, and shuts down.

## Constructing a vector

The *ufovectors* package provides constructors for various types of vectors:

* `ufo_integer_bin (path)`
* `ufo_numeric_bin (path)`
* `ufo_logical_bin (path)`
* `ufo_complex_bin (path)`
* `ufo_raw_bin (path)`

Each of these functions requires a path to a binary file. Our example binary
file at path `../data/100K_seq_int.bin` contains 100,000 32-bit consecutive
little-endian-encoded integer values requiring around 25MB of memory , ie:

```{bash}
hexdump -x ../data/10M_seq_int.bin | head
```

Construct a UFO using the integer version constructor. 

```{r}
bin_vector <- ufo_integer_bin("../data/10M_seq_int.bin", min_load_count=1000)
```

This reads the data from the binary file and interprets it as an integer vector
containing the values `0:100000`.

When we execute this function the R interpreter asks the UFO framework to
allocate some memory using a custom allocator that will be used to store a
vector. However, instead of allocating any of the vector within the process
memory, UFOs are created as a memory mapped area of memory without a mapping to
any physical memory resources, so without contents. This means that initially
the vectors appears to the R runtime as a memory object taking up 40MB of
memory, whereas, the actual physical meomry used is close to zero.

## Accessing the vector

Whenever that memory is accesed, the operating system passes on a request to the
UFO system to allocate and populate a chunk within the mapped memory area. When
that happens, the UFO framework calls a populate function, which, in this case,
will load the data from the binary file. We set the chunk size to `1000`
elements via the `min_load_count` parameter. If not set, the default is as many
elements as fit within 1MB of memory.

Access elements of the vector as follows.
```{r ufovectors-poke-int-vector}
bin_vector[1:10]
```

Once we access an element, the framework allocates and populates a chunk-sized of and
asks its source to populate it. Since the source is a binary file, a chunk of
the file is read into memory. The bytes read from disk are interpreted as 4-byte integer values.

If we access some more elements again, this data is actually in memory and no
more loading takes place. Accessing values from outside the loaded chunk will
cause more chunks to be loaded.

# GC and writeback

When the memory used by materialized chunks reaches the high water mark, the UFO
framework garbage collects least recently loaded chunks until the low water mark
is reached. We show this in `../experiments/membench.Rmd`.

The behavior can be triggered with a scan. We will access every 100,000th elemnt
of the vector.

```{r}
bin_vector[seq.int(1, length(bin_vector), 100000)]
```

As more new chunks are unloaded, old chunks are unbloaded, to maintain bounded
memory usage. When these chunks are accessed again, they are loaded from the
original file.

Vectors can be written to.

```{r}
f <- function() {
  x <- ufo_integer_bin("../data/10M_seq_int.bin", min_load_count=1000)
  x[1] <- as.integer(15)
}
# Note: We have to wrap this in a function to prevent rstudio from grabbing 
# references to the UFOs and forcing them to be copied on write. This does not 
# happen in ordinary R programs.
f()
```

This chunk is now dirty. If it is subsequently unloaded due to memory pressure,
its value will be stored internally on disk by the UFO framework and
subsequently always read from that source instead. 

It will not be forgotten though. We perform a reverse scan over the vector to
demonstrate that the value at index 1 is the same (index 100 in the reversed
vector).

```{r}
f <- function() {
  x <- ufo_integer_bin("../data/10M_seq_int.bin", min_load_count=1000)
  x[1] <- as.integer(15)
  x[rev(seq.int(1, length(x), 100000))]
}
# Note: We have to wrap this in a function to prevent rstudio from grabbing 
# references to the UFOs and forcing them to be copied on write. This does not 
# happen in ordinary R programs.
f()
```

In addition, these file-backed UFOs have writeback. This means that the modified
values are written back to the originalbinary file. This happens when a dirty
chunk is unloaded.

We remove the vector and turn on R's garbage collector to remove all chunks from
memory, forcing the writeback fucntion to fire.

```{r}
# rm(x)
gc()
```

At this point, when inspected (e.g. with `hexedit`) the file will contain a 15
in place of a 0 integer at the initial position.

```{bash}
hexdump -x ../data/10M_seq_int.bin | head
```

Let us restore the vector to the original state.

```{r}
cleanup <- function() {
  x <- ufo_integer_bin("../data/10M_seq_int.bin", min_load_count=1000)
  x[1] <- as.integer(0)
}
cleanup()
gc()
```

## Operators

UFOs attempt to be feature complete and as transparent as possible. A typical
use of vectors involves setting and getting values form them as well as
performing vectorized operations.

Here we create a vector representing a sequence of integers between 1 and
100,000 and adding one to each element:

```{r}
s <- ufo_integer_seq(1, 100000);
is_ufo(s)
```

```{r}
s <- s + 1
is_ufo(s)
```

This will cause R to copy the contents of the UFO over to a new vector, which
will not be a UFO. This can cause the memory to blow out, if the UFO was large.
To prevent this, the UFO R vectors library provides a suite of operators that
return the results of operations on UFO vectors as file-backed vectors. The
contents of these vectors will be transparently stored in persistent memory
mapped storage. This will cause an increase in disk usage while the UFO is live,
but will prevent process memory from being blown out. The disk space will be
reclaimed when the UFO is garbage collected.

To turn on custom UFO operators, load the library with the option `add_class`.

```{r}
# Unload first, otherwise options will not be updated.
detach("package:ufovectors", unload=TRUE)

# Load the library
options(ufos.high_water_mark_mb=2, ufos.low_water_mark_mb=1)
options(ufovectors.add_class = TRUE)
library(ufovectors)
```

First of all, this adds a class attribute to all UFO vectors: 

```{r}
s <- ufo_integer_seq(1, 100000);
is_ufo(s)
class(s)
```

The now class-conscious UFO can will perform S3 generic calls on operators,
redirecting the call to a set of useful operators that return
larger-than-memory--capable UFOs.

```{r}
s <- s + 1
is_ufo(s)
```

If adding a class to the UFO breaks the transparency too much, UFOs can also be
made to override default operators instead of being plugged into the S3 system.

```r
options(ufovectors.overload_operators = TRUE)
```

While we attempt to mitigate any potential problems, this approach is more
dangerous than the S3 approach, since the override will apply to all operators
and may lead to unexpected problems down the line for other objects.

# Guided tour of UFO vector types

There are a few examples of UFO vectors that represent proof of concept
implementations of a number of common use cases for larger-than-memory vectors.

## BZip2 file-backed vectors

These vectors in essence work just like the file-backed vectors we showed above.
However, the vectors are compressed on disk, and decompressed on-the-fly. This
shows that UFOs can expand on the functionality of frameworks using `mmap`
without the user fault mechanism as a backend by applying a transparent
transformation to the data.

```{r}
bz2_vector <- ufo_integer_bz2("../data/10M_seq_int.bin.bz2", min_load_count=1000)
bz2_vector[1:10]
```

We do not provide a writeback function, since after a block were decompressed
and re-compressed it could change size and require realigning all following
blocks in a file.

## CSV file-backed vectors

This backend parses a CSV file and extracts each column into a separate R
vector. The constructor performs an initial scan of the CSV file discovering
columns and inductively determining their types and lengths. The constructor
also makes an index of offsets that helps find clean starting points for parsing
individual chunks, without the need to parse from the start every time. 

The data contains an example CSV file containing four columns:

```
integer,logical,numeric,string
0,      TRUE,   0,      A
1,      FALSE,  .20,    B
2,      NULL,   .40,    C
3,      TRUE,   .60,    D
4,      FALSE   .80,    E
5,      NULL,   1.00,   F
```

The CSV file is read into a series of vectors on the fly once the UFO is created
via a constructor.

```{r}
csv <- ufo_csv("../data/example.csv")
```

 The vectors representing columns are parts of a standard data.frame, and can be accessed via typical R methods.

```{r}
csv$integer[1:10]
```

While its technically possible to implement a writeback function for CSVs, we
currently do not do so, since CSV is a format that is particularly hostile
format to writes in the middle of the file.

## Row bind

The backend produces a logical concatenation two or more other vectors without
creating a copy. The data is looked up in the parent vectors when it is accessed
in the bound vector. The constructor accepts a lists of vectors to bind into a
single vector. The populate function then calculates the mapping between the
indices of the bound vector and the individual elements within the parent vector
and retrieves the appropriate values, including on-the-fly value coersion into a
supertype.

```{r}
vector_1 <- as.integer(1:10)
vector_2 <- as.numeric(1:10) + 0.5
vector_3 <- c(T,F,NA)

vector_123 <- ufo_bind(vector_1, vector_2, vector_3)
vector_123[1:23]
```

We do not provide write-back.

## PostgreSQL

This backend retrieves data directly from a database. It represents a use of
programmable memory objects as database view or a form of object relational
mapping.  The constructor gathers parameters for a database connection and a
query. The populate function executes a transaction retrieving a chunk-sized
portion of the data, parsing the data, and writing it into the vector.

We provide a sample database `ufos` containing a table called `example`:

```{bash}
echo "select * from example where id <= 10" | psql -U $USER -d ufos
```

We can create UFO vectors accessing contents of individual table columns.

```{r}
db_connection_string <- paste0("dbname = ufos user = ", Sys.getenv("USER"))
psql_numeric <- ufo_psql(db_connection_string, table="example", column="numeric")
psql_numeric[1:10]
```

The writeback function updates the contents of the database:

```{r}
f <- function() {
  psql_numeric <- ufo_psql(db_connection_string, table="example", column="numeric")
  psql_numeric[1] <- psql_numeric[1] + 1
}
# Note: We have to wrap this in a function to prevent rstudio from grabbing 
# references to the UFOs and forcing them to be copied on write. This does not 
# happen in ordinary R programs.
f()
```

We remove the vector and turn on R's garbage collector to remove all chunks from
memory, forcing the writeback fucntion to fire.

```{r}
rm(psql_numeric)
gc()
```

Database contents after update:

```{bash}
echo "select * from example where id <= 10" | psql -U $USER -d ufos
```

We currently do not support string columns, because of a fight we are having
with the runtime's allocator.

## Sequences

The from-to-by sequence implementation produces a from-to-by sequences as an
example of a UFO generating values from a sequence or formula rather than from
an external source. The from-to-by sequence is a common type of vector in use in
R programs. It is a vector containing a sequence of integers in a specific range
with a specific interval between the elements. 

```{r}
s <- ufo_integer_seq(from=1, to=10000000, by=2); # a very large sequence
```

## Write protect

This implementation creates a write-protected view of another vector.
The vector contains all the data of the original vector without copying it. It
is set to be write protected, to prevent any writes being executed on the
vector, or set to private, to keep the updates from permeating back into the
original vector. The constructor collects a pointer to an original vector, and
the populate function retrieves data from it.

```{r}
original <- seq(1,1000,3)
protected <- ufo_write_protect(original)
protected[1:10]
```

The new vector is a "reference" of the original but can be modified, without impacting the original. 

When new data is written into the new vector, the old vector is not copied in
its entirety, as would normally be the case with R's copy-on-write semantics.
Instead, only a small chunk is copied over, and this chunk becomes transparently
file-backed to consrve memory.

```{r}
protected[1:10] <- 10:1
protected[1:10]
original[1:10]
```

## Empty vectors

Finally, this backend provides temporary larger-than-memory vectors.
They can be used by application programmers whenever a large R vector must be
created to store the results of computation. As more chunks are written, more
chunks dematerialize, and more disk is used, but the process does not run out of
memory. The constructor is minimal and merely collects parameters of the vector
such as length, type, and initial value. 

```{r}
empty_vector <- ufo_integer(100);
empty_vector
```

The vector can be written into.

```{r}
empty_vector[1:10] <- 42
```

There is no writeback function.