---
title: "UFO Benchmarks"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

~~~ 
Note:

This is an R markdown document. It is like a markdown document, but has
executable snippets of code in in R and bash. It can be viewed using vscode or
RStudio, e.g.:

    rstudio benchmark.Rmd

There, each chunk can be executed via a button over the code. All paths in this
file are relative to the location of the file, so set the session's working
directory to the source file in rstudio before running the snippets:

    Session -> Set working directory -> To source file location

The file can also be executed (in whole) with R using the following bash
snippet:*
 
    Rscript -e "rmarkdown::render('benchmark.Rmd')"

This will compile the document into HTML while executing all the code snippets
in the document. 
~~~

# Introduction

This notebook generates Fig. 12 from the paper. The figure consists of 6 graphs 
which this notebook generates as 6 separate PDFs in subdirectory `fig`. There are 
3 graphs for each  benchmarked back end: creation, sum, and loop. Description from 
the paper follows:

~~~

The experiment benchmark sthe performance of some of the UFO backends. We compare their
behavior to standard R objects, as well as ALTREP vectors. 

The benchmark uses two back-ends. The binary file backend reads 4-byte integers from 
a binary file and has a relatively high overhead when retrieving a single value, which is
amortized when reading entire regions at once. The sequence backend
generates a from-to-by sequence, and computing an individual element of the
sequence is cheap. Both of these backends are implemented in both ALTREP and
UFOs. Pains were taken to ensure that both implementations are equivilent, using
the same system calls and techniques and sharing as much code as possible.

We measured the time it takes to create a 1GB vector (1K iterations), calculate
the sum of its contents (1K iterations), and execute an identity function on
each of its elements (10 iterations).

We test UFOs in two modes: read/write mode and read-only
mode. Read-only mode does not persist changes done to UFOs, which removes the
need to calculate hashes of segment contents when loading and unloading them.

~~~

# Setup

Load libraries used to process data in this document:

```{r}
library(ufovectors)
library(ufoaltrep)
library(microbenchmark)
library(ggplot2)
library(scales)
library(dplyr)
library(readr)
```

Create a few helper functions and do basic settings:

```{r}
plot_to_pdf_and_here <- function(path, data) {
  pdf(paste0("figs/",path), width = 4, height = 4)
  plot(data)
  dev.off()  
  plot(data)
}

dir.create("figs")

some_function <- function(x) x

ufo_set_debug_mode(F)
options(scipen=999) # effectively turns off scientific notation

SMALL_ITER=2
BIG_ITER=10
```

## Generating files

The benchmark uses some on-disk data. We generate a file containing 250M 
random integers (1M random integers repeated 250 times). 

(Other configurations can be uncommented to created files containing sequences 
or a single repeated value.)

```{r}
generate_binary_file <- function(path, range, repeats) {
  sum <- 0
  minimum <- integer(0)
  maximum <- integer(0)
  size <- length(range) * repeats
  f <- file(path, "wb")
  for (i in 0:repeats) {
    writeBin(range, f)
    sum <- sum + sum(range, na.rm = TRUE)
    minimum <- min(minimum, range, na.rm = TRUE)
    maximum <- max(maximum, range, na.rm = TRUE)
  }
  close(f)
  list(path=path, sum=sum, size=size, min=minimum, max=maximum)
}

stats <- list(
  #stats_1s = generate_binary_file("250M_1s.bin", as.integer(rep(1, 1000000)), 250), #126MB
  #stats_seq_ints = generate_binary_file("250M_seq_ints.bin", 1:1000000, 250),
  stats_rand_ints = generate_binary_file("../data/250M_rand_ints.bin", sample(1000000, replace=T), 250)
)

size <- 250 * 1000 * 1000
```

# File-backed vectors

In this section we generate the three graphs for the file-backed vector implementation: UFO and ALTREP. 
UFO vectors are set to read only or read/write. They read data from the binary file created above. 

### Creation

This snippet benchmarks vector creation. It iterates 1000 times. It takes seconds to execute.

```{r}
result <- microbenchmark(
  "UFO"    = { ufo_integer_bin(stats$stats_rand_ints$path, read_only=FALSE) },
  "UFO/RO" = { ufo_integer_bin(stats$stats_rand_ints$path, read_only=TRUE) },
  "ALTREP" = { altrep_ufo_integer_bin(stats$stats_rand_ints$path) },
  times = BIG_ITER
)

plot <- autoplot(result) +
  ggtitle("File-backed/250M rand ints/create")+
  theme_minimal() 

plot_to_pdf_and_here("250M_rand_bin_create.pdf", plot)
```

### Sum

This snippet benchmarks calculating a sum over the entire vector. It iterates 1000 times. It takes minutes to execute.

```{r}
ufo <- ufo_integer_bin(stats$stats_rand_ints$path, read_only=FALSE)
ufo.ro <- ufo_integer_bin(stats$stats_rand_ints$path, read_only=TRUE)
altrep <- altrep_ufo_integer_bin(stats$stats_rand_ints$path)

result <- microbenchmark(
  "UFO"    = { sum(ufo) },
  "UFO/RO" = { sum(ufo.ro) },
  "ALTREP" = { sum(altrep) },
  times = BIG_ITER
)

plot <- autoplot(result) +
  ggtitle("File-backed/250M rand ints/sum")+
  theme_minimal() 

plot_to_pdf_and_here("250M_rand_bin_sum.pdf", plot)
```


### For loop

This snippet benchmarks vector creation. It iterates 10 times. It can take hours to execute.

```{r}
ufo <- ufo_integer_bin(stats$stats_rand_ints$path, read_only=FALSE)
ufo.ro <- ufo_integer_bin(stats$stats_rand_ints$path, read_only=TRUE)
altrep <- altrep_ufo_integer_bin(stats$stats_rand_ints$path)

result <- microbenchmark(
  "UFO"    = { for (e in ufo) some_function(e) },
  "UFO/RO" = { for (e in ufo) some_function(e) },
  "ALTREP" = { for (e in altrep) some_function(e) },
  times = SMALL_ITER
)

plot <- autoplot(result) +
  ggtitle("File-backed/250M rand ints/loop")+
  theme_minimal() 

plot_to_pdf_and_here("250M_rand_bin_loop.pdf", plot)
```

# Sequences

In this section we generate the three graphs for the sequence implementation: UFO and ALTREP. 
UFO vectors are set to read only or read/write. They read data from the binary file created above. 

### Creation

This snippet benchmarks vector creation. It iterates 1000 times. It takes seconds to execute.

```{r}
result <- microbenchmark(
  "UFO"      = { ufo_integer_seq(1, size, 1, read_only = FALSE) },
  "UFO/RO"   = { ufo_integer_seq(1, size, 1, read_only = TRUE) },
  "ALTREP"   = { altrep_integer_seq(1, size, 1) },
  "R vector" = { as.integer(c(1:size)) },
  times = BIG_ITER
)

plot <- autoplot(result) +
  ggtitle("Sequence/250M ints/create") +
  theme_minimal() 

plot_to_pdf_and_here("250M_seq_create.pdf", plot)
```

### Sum

This snippet benchmarks vector creation. It iterates 1000 times. It takes seconds to execute.

```r
ufo <- ufo_integer_seq(1, size, 1, read_only = FALSE)
ufo.ro <- ufo_integer_seq(1, size, 1, read_only = TRUE)
altrep <- altrep_integer_seq(1, size, 1)
vec.std <- as.integer(c(1:size))

result <- microbenchmark(
  "UFO"      = { sum(ufo) },
  "UFO/RO"   = { sum(ufo.ro) },
  "ALTREP"   = { sum(altrep) },
  "R vector" = { sum(vec.std) },
  times =  BIG_ITER
)

plot <- autoplot(result) +
  ggtitle("Sequence/250M ints/sum")+
  theme_minimal() 

plot_to_pdf_and_here("250M_seq_sum.pdf", plot)
```

### For loop

This snippet benchmarks vector creation. It iterates 10 times. It can take hours to execute.

```{r}
ufo <- ufo_integer_seq(1, size, 1, read_only = FALSE)
ufo.ro <- ufo_integer_seq(1, size, 1, read_only = TRUE)
altrep <- altrep_integer_seq(1, size, 1)
vec.std <- as.integer(c(1:size))

result <- microbenchmark(
  "UFO"      = { for (e in ufo) some_function(e) },
  "UFO/RO"   = { for (e in ufo.ro) some_function(e) },
  "ALTREP"   = { for (e in altrep) some_function(e) },
  "R vector" = { for (e in vec.std) some_function(e) },
  times = SMALL_ITER
)

plot <- autoplot(result) +
  ggtitle("Sequence/250M ints/loop")+
  theme_minimal() 

plot_to_pdf_and_here("250M_seq_loop.pdf", plot)
```

# Discussion

The discussion of the experimental results from the paper:

~~~

We show the results of the evaluation in Fig. 12. Each plot
shows the results for either the creation, sum, or loop microbenchmark. The top
row shows results for the file-backed vectors, the bottom one for sequences. The
Y-axis always shows vector implementations and the X-axis shows execution time.
The results are shown as a violin plots showing the relative distribution of
execution times over multiple iterations.

We observe that UFOs and ALTREP have similar performance for vector creation
and the execution time is negligibly small for both frameworks, with some
outliers we attribute to initialization and garbage collection. The startup
time is higher for R vectors implementing a sequence, because the vector must
populated up front, as opposed to UFOs and ALTREP, which calculate these values
on demand. This initialization cost for standard vectors could eventually be
amortized over multiple passes over the vector.

Sums over sequences yield similar performance for all frameworks. Sequences,
being lightweight to calculate, would highlight any difference in implementation
overhead, though little is observed. 
%%
For file-backed vectors UFOs and ALTREP also perform similarly.  The R runtime
calculates the sum of a vector using a fast arithmetic function. This function
cooperates with ALTREP to chunk the vector into regions, which allows ALTREP
to amortize the overhead of preparing a file for reading and seeking.  While
the R runtime does not similarly chunk the execution for UFOs, the UFO
framework makes sure to read no less than 1MB of elements at-a-time and cache
data, yielding a similar amortization. Thus, the performance for both
frameworks is similar.  When the hashing mechanism is turned off for read-only
UFO vectors, a significant overhead cost is removed for UFOs, yielding a small,
but visible improvement in performance.

An importance difference in performance between UFOs and ALTREP stems from the
fact that ALTREP performs dynamic dispatch whenever values are accessed, be it
a region or a single value. The R runtime attempts to turn individual value
accesses into region accesses for ALTREP, but this can only works for specific
operations. When the loop benchmark executes, it always executes a function on
a single value from a vector, leading to repeated dispatch in ALTREP, and so,
deteriorates performance significantly. UFOs also have set-up costs relating to
loading data for an accessed value, however these costs are always amortized by
loading an entire segment into memory. This gives UFOs an advantage over
ALTREP's dispatch and produces performance close to ordinary vectors when
consecutive elements are accessed. However, this approach is costly if
the access pattern is spread out, causing the UFO to load and
unload a segment for each single value read.

~~~