---
title: "Demo"
output: html_document
---

~~~
Note:

This is an R markdown document. It is like a markdown document, but has
executable snippets of code in in R and bash. It can be viewed using vscode or
RStudio, e.g.:

    rstudio ufo-r-vectors.Rmd

There, each chunk can be executed via a button over the code. All paths in this
file are relative to the location of the file, so set the session's working
directory to the source file in rstudio before running the snippets:

    Session -> Set working directory -> To source file location

The file can also be executed (in whole) with R using the following bash
snippet:*
 
    Rscript -e "rmarkdown::render('ufo-r-vectors.Rmd')"

This will compile the document into HTML while executing all the code snippets
in the document. 
~~~

This is a quick demonstration of UFOs and how they perform vs R vectors
 as well as a comparable AltRep implementation of file backed arrays

It is meant to generate similar graphs to those presented in our paper
 but be much faster to run.

# Setup

First we will set up some preliminaries.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


plot_to_pdf_and_here <- function(path, data) {
  pdf(path, width = 4, height = 4)
  plot(data)
  dev.off()
  
  plot(data)
}
```

```{r message = FALSE, warning = FALSE }
# Load the library
library(ufovectors)

# and also the other libraries
library(ufoaltrep)
library(microbenchmark)
library(ggplot2)
library(scales)
library(dplyr)
library(readr)
```

```{R echo = FALSE  }
ufo_set_debug_mode(F)

options(scipen=999) # effectively turns off scientific notation
```

### Data

And setup a file with some dummy-data to use in tests

```{r generate, message = FALSE, warning = FALSE }
generate_binary_file <- function(path, range, repeats) {
  sum <- 0
  minimum <- integer(0)
  maximum <- integer(0)
  size <- length(range) * repeats
  f <- file(path, "wb")
  for (i in 0:repeats) {
    writeBin(range, f)
    sum <- sum + sum(range, na.rm = TRUE)
    minimum <- min(minimum, range, na.rm = TRUE)
    maximum <- max(maximum, range, na.rm = TRUE)
  }
  close(f)
  list(path=path, sum=sum, size=size, min=minimum, max=maximum)
}

stats <- list(
  # just a File Backed list of numbers 1-1M repeated 32 times to stand in for more interesting data
  stats_seq_ints = generate_binary_file("data/demo_32M_seq_ints.bin", 1:1000000, 32)
)
```

### Constants for sizes and sampling

To keep the runtimes down and not be too taxing on your computer we will only
 use 128MB of data. When doing element-wise operations we will also be sub-sampling
 that data only 100,000 times. This allows more runs and nicer more accurate charts.

```{R  message = FALSE, warning = FALSE }
# 32 million elements in the test sets, a nice round number
# these are 32-bit (4 byte) ints so that is 128 megs
size <- 32 * 1000 * 1000
# some tests would run a long time if all elements were used, take a sample
sample_size <- 100 * 1000
set.seed(14) # Initialize the RNG with a nothing up my sleeve number

# a function that does nothing, but forces R to call it
some_function <- function(x) x
```

## File-Backed instances

First we'll compare to AltRep using file-backed arrays.
Both UFO and AltRep implementations are using the same core function to load data
 from the file.

### Sum

Doing a straight sum over all elements is a fast-path for R and will make good use
 of AltRep's ability to load chunks of data in one call.

```{r fileBacked_sum, cache=T, fig.height=3, fig.width=3, message = FALSE, warning = FALSE }
(function(){
ufo <- ufo_integer_bin(stats$stats_seq_ints$path, read_only=FALSE)
ufo.ro <- ufo_integer_bin(stats$stats_seq_ints$path, read_only=TRUE)
altrep <- altrep_ufo_integer_bin(stats$stats_seq_ints$path)

result <- microbenchmark(
  "UFO" = { sum(ufo) },
  "UFO/RO" = { sum(ufo.ro) },
  "ALTREP" = { sum(altrep) },
  times = 50L
)

autoplot(result) +
  ggtitle("File-backed/32M seq/sum")+
  theme_minimal() 
})()
```

Performance favors alt-rep on sums,
  though both a good job with this because of loading large chunks

### Random access

In Random access there are two forces at work.
One is that for UFOs we must load whole chunks to access even one element
The other is that AltRep does a lot of dispatch, which is heavier?

```{r fileBacked_random, cache=T, fig.height=3, fig.width=3, message = FALSE, warning = FALSE }
(function(){
ufo <- ufo_integer_bin(stats$stats_seq_ints$path, read_only=FALSE)
ufo.ro <- ufo_integer_bin(stats$stats_seq_ints$path, read_only=TRUE)
altrep <- altrep_ufo_integer_bin(stats$stats_seq_ints$path)

result <- microbenchmark(
  # R is rather slow with element-wise access
  #  use the smaller sample size and run it a fewer times
  "UFO" = { 
    for (i in sample(size, sample_size)) some_function(ufo[i])
  },
  "UFO/RO" = {
    for (i in sample(size, sample_size)) some_function(ufo.ro[i])
  },
  "ALTREP" = {
    for (i in sample(size, sample_size)) some_function(altrep[i])
  },
  times = 10L
)

autoplot(result) +
  ggtitle("File-backed/32M seq/random access")+
  theme_minimal() 
})()
```

AltRep is much heavier than the UFOs in this case.
Since we are reading randomly from the vectors AltRep loads elements singly while UFOs still loads chunks.

## In-memory Arrays

Loading things from disk is perhaps not always a good comparison, however.
There are cases where the object would be held in memory but can't be because it is too large.
So now we will compare against a normal in-memory R vector

### Sum
```{R inMemory_sum, cache=T, fig.height=3, fig.width=3, message = FALSE, warning = FALSE}
(function(){
ufo <- ufo_integer_seq(1, size, 1, read_only = FALSE)
ufo.ro <- ufo_integer_seq(1, size, 1, read_only = TRUE)
vec <- as.integer(c(1:size))

result <- microbenchmark(
  # R is rather slow with element-wise access
  #  use the smaller sample size and run it a fewer times
  "UFO" = { sum(ufo) },
  "UFO/RO" = { sum(ufo.ro) },
  "R Vector" = { sum(vec) },
  times = 100L
)

autoplot(result) +
  ggtitle("In Memory/32M seq/sum")+
  theme_minimal() 
})()
```

UFOs are very comparable to plain R vectors.
They have a longer tail of long running instances for things like Kernel thread scheduling,
 but most runs are in the same range as plain vectors

### Random Access

Random access is harder on R, so this test is slower than sums, like above.

```{R inMemory_random, cache=T, fig.height=3, fig.width=3, message = FALSE, warning = FALSE }
(function(){
ufo <- ufo_integer_seq(1, size, 1, read_only = FALSE)
ufo.ro <- ufo_integer_seq(1, size, 1, read_only = TRUE)
vec <- as.integer(c(1:size))

result <- microbenchmark(
  # R is rather slow with element-wise access
  #  use the smaller sample size and run it a fewer times
  "UFO" = { 
    for (i in sample(size, sample_size)) some_function(ufo[i])
  },
  "UFO/RO" = {
    for (i in sample(size, sample_size)) some_function(ufo.ro[i])
  },
  "R Vector" = {
    for (i in sample(size, sample_size)) some_function(vec[i])
  },
  times = 10L
)

autoplot(result) +
  ggtitle("In Memory/32M seq/random access")+
  theme_minimal() 
})()
```

Random access tells a similar story as sums when comparing to R vectors

Comparing to R Vectors shows that UFOs have very little overhead even compared to
  completly native structures